---
title: 'data.table: Concatenation'
author: "D.McCabe"
date: "23 March 2018"
output: html_document
---

# Introduction

The following is a brief introduction to combining data.tables.  In the examples I'll use a subset from a data.gov.uk dataset. The values in the table correspond to the proportion of large fish (equal to or larger than 40cm), by weight, in the northern North Sea.
```{r setup}
library(data.table)

dt<-fread(
 input="http://data.defra.gov.uk/env/wdfg21_mar_eco_200904.csv"
 ,header=T
 ,skip=2
 ,nrows=27
 ,na.strings="-"
 ,drop=c(4,5,6)
)
dt<-dt[complete.cases(dt)]
head(dt)
```


# Combining Columns

To illustrate column merging we'll first break the data up by fish type then later recombine it to illustrate the merge
```{r column_splitting}
northern<-dt[,.(Year,`Northern North Sea`)]
southern<-dt[,.(Year,`Southern North Sea`)]
```

Next we'll recombine the data.tables using the Year column as a key to match rows.
```{r column_merging}
  colmerge<-merge(
    x = northern,
    y = southern,
    by = c('Year')
  )
  head(colmerge)
```


# Combining Rows

To illustrate row merging we'll first break the data up by year before recombine it.
```{r row_splitting}
pre2000<-dt[Year<2000]
year2000<-dt[Year==2000]
post2000<-dt[Year>2000]
```

To recombine these data.tables we can group them in a list then use the rbindlist to join them together.
```{r row_merging}
rowmerge<-rbindlist(list(post2000,pre2000,year2000),fill=T)
rowmerge<-rowmerge[order(Year)] # ensure correct row order if desired
print(rowmerge)
```
