---
title: "Hierachical Clustering"
author: "D.McCabe"
date: "16 August 2017"
output: html_document
---

- An agglomerative approach
-- Find closest two points
-- Pair them together
-- Repeat...

- Requires
- - Defined distance metric
- - - [geometric](https://en.wikipedia.org/wiki/Euclidean_distance)
- - - [similarity](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient)
- - - [binary](https://en.wikipedia.org/wiki/Taxicab_geometry)
- - A merging approach
- - - Averaging - take the midpoint/centroid
- - - Complete - take the distance between the farthest pair in each cluster
- Produces
-- A tree showing the mutual proximity of all the points

```{r}
set.seed(71675)
par(mar=c(0,0,0,0))
x<-rnorm(12,mean=rep(1:3,each = 4), sd = 0.2)
y<-rnorm(12,mean=rep(c(1,2,1), each = 4), sd = 0.2)
plot(x,y, col="blue", pch=19,cex=2)
text(x+0.05, y+0.05, labels = as.character(1:12))
```

##### Dist
```{r}
library(data.table)
dt<-data.table(x=x,y=y)
dist(dt)
```

##### Pretty Dendrograms
```{r}
distxy<- dist(dt)
hClustering<-hclust(distxy)

# Function to color branches
colbranches <- function(node, col){
  
  # Find the attributes of current node
  a <- attributes(node)
  
  # Color edges with requested color
  attr(node, "edgePar") <- c(a$edgePar, list(col=col, lwd=2))
  
  # Return the node!
  node
}

# get the dendrogram object
dendro <- as.dendrogram(hClustering)

dendro[[2]][[1]] = dendrapply(dendro[[2]][[1]], colbranches, "orange")
dendro[[2]][[2]] = dendrapply(dendro[[2]][[2]], colbranches, "red")
dendro[[1]] = dendrapply(dendro[[1]], colbranches, "blue")

plot(dendro)
```

##### Heatmap

```{r}
dataMatrix<-as.matrix(dt)[sample(1:12),]
heatmap(dataMatrix)
```
