---
title: Study To Establishing Which Extreme U.S. Weather Events Have BeenResponsible For The Highest Humanitarian and Economic Loses
author: "D.McCabe"
date: "27 September 2017"
output: html_document
---


## Synopsis

[survivorship bias](https://en.wikipedia.org/wiki/Survivorship_bias)
[texas sharpshooter](https://en.wikipedia.org/wiki/Texas_sharpshooter_fallacy)

summarize analysis in less than 10 sentences
Three part analysis
- type(s) of event that cause the greatest harm to human health (how that was reached)
- type(s) of event that cause the greatest economic losses (how that was reached)
- observe correlations??
- what do the results suggest?
- where there any issues with the quality of the data? Any further investigations before conclusive comments?



## Data Processing
- source of data, need to seek from Url unzip ...
- the data was procured from ... on ...
```{r,echo=FALSE}
library(tidyr)
library(ggplot2)
library(data.table)
library(scales)

# perform data munging
source("data_munging.R")
```

Occurance
```{r}

# display top 10s
{
  # Top 10 Fatalities
  fatalities_by_weather_event<-dt[
    as.integer(year(BGN_TIME))>1994,
    .(
      TOT_FATALITIES=sum(FATALITIES),
      AVE_FATALITIES=mean(FATALITIES),
      SD_FATALITIES=sd(FATALITIES)
    ),
    EVTYPE
  ][order(-TOT_FATALITIES,-AVE_FATALITIES)]
  
  head(fatalities_by_weather_event,3)
  
  # Top 10 Injuries
  injuries_by_weather_event<-dt[
    as.integer(year(BGN_TIME))>1994,
    .(
      TOT_INJURIES=sum(INJURIES),
      AVE_INJURIES=mean(INJURIES),
      SD_INJURIES=sd(INJURIES)
    ),
    EVTYPE
  ][order(-TOT_INJURIES,-AVE_INJURIES)]
  
  head(injuries_by_weather_event,3)
}

```


```{r}

# display top 10s
{
  # Top 10 Fatalities
  fatalities_by_weather_event<-dt[
    ,
    .(
      TOT_FATALITIES=sum(FATALITIES),
      AVE_FATALITIES=mean(FATALITIES),
      SD_FATALITIES=sd(FATALITIES)
    ),
    EVTYPE
  ][order(-TOT_FATALITIES,-AVE_FATALITIES)]
  
  head(fatalities_by_weather_event,3)
  
  # Top 10 Injuries
  injuries_by_weather_event<-dt[
    ,
    .(
      TOT_INJURIES=sum(INJURIES),
      AVE_INJURIES=mean(INJURIES),
      SD_INJURIES=sd(INJURIES)
    ),
    EVTYPE
  ][order(-TOT_INJURIES,-AVE_INJURIES)]
  
  head(injuries_by_weather_event,3)
}

```

## Results
- 3 graphs or less
